---
title: "6. Analysis_pipeline_All"
output: html_document
date: "2025-05-20"
---

# 0. Setup and Library Imports
```{r}
rm(list=ls())
gc()
```

```{r}
library(devtools)
library(plyr)
library(dplyr)
library(monocle3)
library(ggplot2)
library(ggridges)
library(ggrepel)
library(ggpubr)
library(gridExtra)
library(pheatmap)
library(tibble)
library(tidyr)
library(parallel)
library(DelayedArray)
library(Matrix)
library(magrittr)
library(purrr)
library(furrr)
library(future)
library(progressr)
library(cli)
```

```{r}
source("/burg/iicd/users/qc2358/Kinase_project/github_repos/sci-plex/bin/cell_cycle.R")
source("/burg/iicd/users/qc2358/Kinase_project/github_repos/sci-plex/bin/dispersions_functions.R")
cc.genes = readRDS("/burg/iicd/users/qc2358/Kinase_project/github_repos/sci-plex/bin/cc.genes.RDS")
```

```{r}
# Define the number of cores to use for parallel processing
n_cores <- 20
message(paste("Using", n_cores, "cores for parallel processing."))
# Adjust this value based on your system memory and needs.
options(future.globals.maxSize = 6 * 1024^3) # 6GB

handlers(handler_progress(
  format = ":spin [:bar] :current/:total (:percent) | ETA: :eta | Processing: :message",
  clear = FALSE
))
# Set up the parallel plan
plan(multisession, workers = n_cores)
```

```{r}
output_folder <- "CRISPRI_Figures/"
base_dir <- "/burg/iicd/users/qc2358/Kinase_project"
output_path <- file.path(base_dir, output_folder)
setwd(base_dir)

cds_folder <- "preprocessed_cds_without_Tcells.rds"
cds_path <- file.path(base_dir, cds_folder)
cds <- readRDS(cds_path)
cds <- cds[, !is.na(colData(cds)$top_sg)]
dim(cds) # 58347 166061
```

# 2. UMAP Visualization
```{r}
expressed_genes <- row.names(fData(cds)[Matrix::rowSums(exprs(cds) > 0) > 500,])

dispersions_test <- estimateDispersionsForCellDataSet(cds,
                                                      min_cells_detected = 1000,
                                                      removeOutliers = FALSE)

disp_table <- dispersionTable(dispersions_test)

disp_table <- disp_table %>%
  mutate(excess_disp = (dispersion_empirical - dispersion_fit) / dispersion_fit)

unsup_clustering_genes <- disp_table %>%
  dplyr::arrange(plyr::desc(excess_disp)) %>%
  dplyr::slice(1:2000) %>%
  pull(gene_id)

print(length(unique(unsup_clustering_genes)))
```

```{r}
cds <- preprocess_cds(cds,
                      num_dim = 15,
                      use_genes = unsup_clustering_genes)

cds <- reduce_dimension(cds, 
                        umap.n_neighbors = 20L,
                        umap.min_dist = 0.1,
                        max_components = 2,
                        #preprocess_method = "PCA",
                        umap.fast_sgd = FALSE,
                        cores = 1)

colData(cds)$UMAP1 <- reducedDims(cds)[["UMAP"]][, 1]
colData(cds)$UMAP2 <- reducedDims(cds)[["UMAP"]][, 2]
```

```{r}
cds <- cluster_cells(cds, reduction_method = "PCA", resolution = 2e-4, random_seed = 2016L, num_iter = 1)
colData(cds)$PCA_Cluster <- clusters(cds, reduction_method = "PCA")
length(unique(colData(cds)$PCA_Cluster))
```

```{r}
plot_cells(cds, color_cells_by = "PCA_Cluster") +
  theme(legend.position = "right",
        legend.key.width = unit(0.2,"line"),
        legend.key.height = unit(0.2,"line"))
ggsave("CRISPRI_Figures/UMAPs/UMAP_PCA_cluster.png", width = 4, height = 4, dpi = 300)

plot_cells(cds, color_cells_by = "dose") +
  theme(legend.position = "right",
        legend.key.width = unit(0.2,"line"),
        legend.key.height = unit(0.2,"line"))
ggsave("CRISPRI_Figures/UMAPs/UMAP_dose.png", width = 4, height = 4, dpi = 300)

colData(cds)$dosage <- as.character(colData(cds)$dose)
dosage_colors<- c('0' = '#A6CEE3','0.25'='#1F78B4','0.5'='#B2DF8A','1' = '#33A02C')
plot_cells(cds, color_cells_by = "dosage") +
  scale_color_manual(values = dosage_colors)+
  theme_void()+
  theme(legend.position = "right",
        legend.key.width = unit(0.2,"line"),
        legend.key.height = unit(0.2,"line"))
ggsave("CRISPRI_Figures/UMAPs/UMAP_dosage.png", width = 4, height = 4, dpi = 300)
```

```{r}
# Plot cells using custom genes and viridis color palette
plot_cells(cds,
           genes = c('SOD2', 'CSF3','CD274','IDO1'),
           show_trajectory_graph = FALSE) +
  viridis::scale_color_viridis(option = "magma") +  # Use the viridis "magma" palette
  theme(legend.position = "right",
        legend.key.width = unit(0.2, "line"),
        legend.key.height = unit(0.2, "line"))
ggsave("CRISPRI_Figures/UMAPs/UMAP_SOD2_CSF3.png", width = 4, height = 4, dpi = 300)
```

```{r}
saveRDS(cds, "cds_2000.rds")
```

# 3. Genes associated with T cell exposure
## 3.1 UMAP of top 30 DEGs
```{r}
Tcell_dose_diff_test = readRDS("Figure1/DEG_testing/NTC_CRISPRI_dose_response_diff_test_20250520_replicates_ori.rds")
Tcell_dose_diff_test <- Tcell_dose_diff_test %>%
  filter(!is.na(p_value)) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))
```

```{r}
Tcell_dose_response_diff_test_top_deg <- Tcell_dose_diff_test[Tcell_dose_diff_test$q_value < 0.05 & Tcell_dose_diff_test$term == "dose",]
Tcell_dose_response_diff_test_top_deg <- Tcell_dose_response_diff_test_top_deg[order(Tcell_dose_response_diff_test_top_deg$q_value),]
```

```{r}
top30 <- head(Tcell_dose_response_diff_test_top_deg, 30)
top_Tcell_dose_gene <- top30[top30$normalized_effect > 1.3,]
plot_cells(cds,
           genes = c(top_Tcell_dose_gene$gene_short_name),
           show_trajectory_graph = FALSE) +
  viridis::scale_color_viridis(option = "magma") +  # Use the viridis "magma" palette
  theme(legend.position = "right",
        legend.key.width = unit(0.2, "line"),
        legend.key.height = unit(0.2, "line"))
ggsave("CRISPRI_Figures/UMAPs/UMAP_top_tcell_dose.png", width = 8, height = 8, dpi = 300)
```

## 3.2 DEG analysis for each kinase in untreated cells
```{r}
kinases <- (unique(colData(cds)$gene_id))
kinases <- kinases[kinases != "NTC"]
kinases <- na.omit(kinases)
```

```{r}
untreated_kinase_cds <- cds[,!is.na(colData(cds)$dose) & !is.na(colData(cds)$gene_id) & colData(cds)$dose == 0]
median(table(colData(untreated_kinase_cds)$gene_id)) # 78
```

```{r}
cli::cli_h1("DEG analysis for each kinase in untreated cells")
message("Processing ", length(kinases), " kinases")

# Wrap the parallel computation with with_progress to enable progress reporting
with_progress({
  # Create a progressor
  p <- progressor(steps = length(kinases))
  
  untreated_genotype_diff_test.list <- furrr::future_map(kinases, function(target) {
    p(message = sprintf("%s", target))
    # Create subset for the current target kinase and NTC
    cds_subset <- untreated_kinase_cds[,colData(untreated_kinase_cds)$gene_id %in% c("NTC", target)]
    colData(cds_subset)$gene_id <- factor(colData(cds_subset)$gene_id, levels = c("NTC", target))
    
    expressed_genes <- row.names(fData(cds_subset)[Matrix::rowSums(exprs(cds_subset) > 0) > dim(cds_subset)[2]*0.05,])
    model_fit <- fit_models(cds_subset[expressed_genes,], model_formula_str = "~gene_id", cores = 1)
    result_table <- coefficient_table(model_fit) %>%
      dplyr::select(-model, -model_summary)
    
    return(result_table)
  }, .options = furrr_options(seed = TRUE)) # seed = TRUE for reproducibility
})
names(untreated_genotype_diff_test.list) <- kinases
untreated_genotype_diff_test.list <- purrr::compact(untreated_genotype_diff_test.list) # Remove NULL entries
```

```{r}
# Unparallelized version
# untreated_genotype_diff_test.list <- list()
# for(target in kinases){
#   cds_subset <- untreated_kinase_cds[,colData(untreated_kinase_cds)$gene_id %in% c("NTC",target)]
#   colData(cds_subset)$gene_id <- factor(colData(cds_subset)$gene_id, levels = c("NTC",target))
#   expressed_genes <- row.names(fData(cds_subset)[Matrix::rowSums(exprs(cds_subset) > 0) > dim(cds_subset)[2]*0.05,])
#   untreated_genotype_diff_test.list[[target]] <- fit_models(cds_subset[expressed_genes,], model_formula_str = "~gene_id")
#   untreated_genotype_diff_test.list[[target]] <- coefficient_table(untreated_genotype_diff_test.list[[target]]) %>%
#     dplyr::select(-model,-model_summary)
    
#   rm(cds_subset)
#   message("Done with ", target)
  
# }
```

```{r}
saveRDS(untreated_genotype_diff_test.list, "CRISPRI_Figures/Diff_test/untreated_genotype_diff_test.list.rds")
```

```{r}
# untreated_genotype_diff_test.list<- readRDS("CRISPRI_Figures/Diff_test/untreated_genotype_diff_test.list.rds")
for(target in names(untreated_genotype_diff_test.list)){
  untreated_genotype_diff_test.list[[target]]$gene_id <- rep(target, nrow(untreated_genotype_diff_test.list[[target]]))
  }

untreated_genotype_diff_results <- do.call(rbind, untreated_genotype_diff_test.list)
untreated_genotype_diff_results$q_value <- p.adjust(untreated_genotype_diff_results$p_value, method = "BH")
```

```{r}
ggplot(untreated_genotype_diff_results %>% filter(term != "(Intercept)"), 
       aes(x = normalized_effect, y = -log10(q_value), color = q_value < 1e-4)) +
  geom_point(size = 0.5, stroke = 0) +
  scale_color_manual(values = c("black", "firebrick2")) +
  facet_wrap(~gene_id == "originalIDTorder") +
  monocle3:::monocle_theme_opts()
ggsave("CRISPRI_Figures/Diff_test/test.png")
```

## 3.3 Generate list of T cell dosage genes controlled by kinases
```{r}
Tcell_dose_diff_genes <- Tcell_dose_diff_test %>% filter(term == "dose", q_value < 0.05) %>% pull(id)
```

```{r}
untreated_genotype_diff_results %>%
  filter(term != "(Intercept)",
         q_value < 0.05,
         abs(normalized_effect) > 1,
         gene_id %in% c("random", "originalIDTorder")) %>%
  group_by(gene_id) %>%
  dplyr::summarise(n = n()) %>%
  arrange(desc(n)) %>%
  print(n = 100)
```

```{r}
untreated_genotype_diff_results %>%
  filter(term != "(Intercept)", q_value < 0.05, abs(normalized_effect) > 1) %>% group_by(gene_id) %>%
  dplyr::summarise(n=n()) %>%
  arrange(desc(n)) %>%
  print(n = 100)
```

```{r}
Tcell_dose_kinase_diff_genes <- untreated_genotype_diff_results %>%
  filter(term != "(Intercept)", q_value < 1e-4, id %in% Tcell_dose_diff_genes) %>% 
  pull(id) %>%
  unique()

length(unique(Tcell_dose_diff_genes)) # 1768
length(Tcell_dose_kinase_diff_genes) # 501
```

## 3.4 DEG analysis for knockdown effect for each treatment condition
```{r}
hashed_cds <- cds[, !is.na(colData(cds)$dose)]
kinase_cds <- hashed_cds[, !is.na(colData(hashed_cds)$gene_id)]
```

```{r}
cli::cli_h1("DEG analysis (T cell dose genes) for each kinase for each treatment condition")

genotype_by_Tcell_dose_diff_test.list <- list()

for(dose in unique(as.character(colData(kinase_cds)$dosage))){
  message(paste("Processing dose:", dose))
  kinase_cds_subset <- kinase_cds[, colData(kinase_cds)$dosage == dose]

  with_progress({
    p_dose <- progressor(steps = length(kinases))
    
    # Parallelize the inner loop over kinases for current dose
    results_for_dose_list <- furrr::future_map(kinases, function(target_kinase) {
      p_dose(message = sprintf("%s", target_kinase)) 
      cds_subset <- kinase_cds_subset[, colData(kinase_cds_subset)$gene_id %in% c("NTC", target_kinase)]
      colData(cds_subset)$gene_id <- factor(colData(cds_subset)$gene_id, levels = c("NTC", target_kinase))
      
      model_fit <- fit_models(cds_subset[Tcell_dose_diff_genes, ], model_formula_str = "~gene_id", cores = 1)
      coef_table <- coefficient_table(model_fit) %>%
        dplyr::select(-model, -model_summary)
      
      return(coef_table)
    }, .options = furrr_options(seed = TRUE))
    
    names(results_for_dose_list) <- kinases
    genotype_by_Tcell_dose_diff_test.list[[as.character(dose)]] <- purrr::compact(results_for_dose_list) # Remove NULL entries
  })
}
```

```{r}
# Unparallelized version
# colData(kinase_cds)$treatment= colData(kinase_cds)$dosage
# for(dose in unique(colData(kinase_cds)$treatment)){
  
#   genotype_by_Tcell_dose_diff_test.list[[dose]] <- list()
  
#   for(target in kinases){
#     cds_subset <- kinase_cds[,colData(kinase_cds)$gene_id %in% c("NTC",target) & colData(kinase_cds)$treatment == dose]
#     colData(cds_subset)$gene_id <- factor(colData(cds_subset)$gene_id, levels = c("NTC",target))
    
#     genotype_by_Tcell_dose_diff_test.list[[dose]][[target]] <- fit_models(cds_subset[Tcell_dose_diff_genes,],
#                                                               model_formula_str = "~gene_id")
#     genotype_by_Tcell_dose_diff_test.list[[dose]][[target]] <- coefficient_table(genotype_by_Tcell_dose_diff_test.list[[dose]][[target]]) %>%
#       dplyr::select(-model,-model_summary)
    
#     rm(cds_subset)
#     message("Done with T cell test for ", dose, " exposed ", target)
    
#   }
# }
```

```{r}
saveRDS(genotype_by_Tcell_dose_diff_test.list, "CRISPRI_Figures/Diff_test/genotype_by_Tcell_dose_diff_test.list.rds")
```

```{r}
#genotype_by_Tcell_dose_diff_test.list <- readRDS("CRISPRI_Figures/Diff_test/genotype_by_Tcell_dose_diff_test.list.rds")

for(dose in names(genotype_by_Tcell_dose_diff_test.list)){
    for(target in kinases){
  genotype_by_Tcell_dose_diff_test.list[[dose]][[target]]$gene_id <- rep(target, nrow(genotype_by_Tcell_dose_diff_test.list[[dose]][[target]]))
  }
}

for(dose in names(genotype_by_Tcell_dose_diff_test.list)){
  
  genotype_by_Tcell_dose_diff_test.list[[dose]] <- do.call("rbind", genotype_by_Tcell_dose_diff_test.list[[dose]])
  genotype_by_Tcell_dose_diff_test.list[[dose]]$treatment <- rep(dose, nrow(genotype_by_Tcell_dose_diff_test.list[[dose]]))
}
```

```{r}
genotype_by_Tcell_dose_diff_test_results <- do.call("rbind", genotype_by_Tcell_dose_diff_test.list)

genotype_by_Tcell_dose_diff_test_results$q_value <- p.adjust(genotype_by_Tcell_dose_diff_test_results$p_value, method = "BH")
```

```{r}
ggplot(genotype_by_Tcell_dose_diff_test_results %>% filter(!(term %in% c("(Intercept)"))),
       aes(x = normalized_effect, y = -log10(q_value), color = q_value < 1e-4 & abs(normalized_effect) > 1)) +
  geom_point(size = 0.5, stroke = 0) +
  facet_wrap(~treatment) +
  monocle3:::monocle_theme_opts() +
  scale_color_manual(values = c("TRUE" = "firebrick2", "FALSE" = "black"))
ggsave("CRISPRI_Figures/Diff_test/test_2.png")
```

```{r}
top_tcell_exposure_perturbed_genes <- genotype_by_Tcell_dose_diff_test_results %>%
  filter(term != "(Intercept)",
         q_value < 0.05,
         abs(normalized_effect) > 1) %>%
  group_by(id) %>%
  summarize(n = n())%>%
  filter(n >= 50) %>%
  pull(id) %>%
  unique()
```

## 3.5 T cell dose genes for each genotype
```{r}
cli::cli_h1("T cell dose genes for each genotype")

with_progress({
  p_genotype <- progressor(steps = length(kinases))

  Tcell_genotype_diff_test.list <- furrr::future_map(kinases, function(target_kinase) {
    p_genotype(message = sprintf("%s", target_kinase))
    cds_subset <- kinase_cds[,colData(kinase_cds)$gene_id %in% c(target_kinase)]
    expressed_genes <- row.names(fData(cds_subset)[Matrix::rowSums(exprs
    (cds_subset) > 0) > dim(cds_subset)[2]*0.05,])

    model_fit <- fit_models(cds_subset[expressed_genes,], model_formula_str = "~dose", cores = 1)
    coef_table <- coefficient_table(model_fit) %>%
      dplyr::select(-model, -model_summary)

    return(coef_table)
  }, .options = furrr_options(seed = TRUE))

  names(Tcell_genotype_diff_test.list) <- kinases
  Tcell_genotype_diff_test.list <- purrr::compact(Tcell_genotype_diff_test.list)
}) # Remove NULL entries
```

```{r}
# Unparallelized version
# Tcell_genotype_diff_test.list <- list()

# for(target in kinases ){
# 	cds_subset <- kinase_cds[,colData(kinase_cds)$gene_id %in% c(target)]
# 	expressed_genes <- row.names(fData(cds_subset)[Matrix::rowSums(exprs(cds_subset) > 0) > dim(cds_subset)[2]*0.05,])
#    Tcell_genotype_diff_test.list[[target]] <- fit_models(cds_subset[expressed_genes,], model_formula_str = "~dose", cores = 1)
#    Tcell_genotype_diff_test.list[[target]] <-coefficient_table(Tcell_genotype_diff_test.list[[target]]) %>%
#     dplyr::select(-model,-model_summary)
#    rm(cds_subset)
#    message("Done with T cell test for", target)
# }
```

```{r}
saveRDS(Tcell_genotype_diff_test.list, "CRISPRI_Figures/Diff_test/Tcell_genotype_diff_test.list.rds")
```

```{r}
# Tcell_genotype_diff_test.list <- readRDS("CRISPRI_Figures/Diff_test/Tcell_genotype_diff_test.list.rds")
for(target in names(Tcell_genotype_diff_test.list)){
  Tcell_genotype_diff_test.list[[target]]$gene_id <- rep(target, nrow(Tcell_genotype_diff_test.list[[target]]))
}
```

```{r}
Tcell_genotype_diff_results <- do.call(rbind, Tcell_genotype_diff_test.list)
Tcell_genotype_diff_results$q_value <- p.adjust(Tcell_genotype_diff_results$p_value, method = "BH")
```

```{r}
ggplot(Tcell_genotype_diff_results %>% filter(!(term %in% c("(Intercept)","log(Tcell_dose + 0.1)"))),
       aes(x = normalized_effect, y = -log10(q_value), color = q_value < 1e-2 & abs(normalized_effect) > 0.5)) +
  geom_point(size = 0.5, stroke = 0) +
  facet_wrap(~gene_id %in% c("random", "originalIDTorder")) +
  monocle3:::monocle_theme_opts() +
  scale_color_manual(values = c("TRUE" = "firebrick2", "FALSE" = "black"))
ggsave("CRISPRI_Figures/Diff_test/test_3.png")
```

# 4. Heatmap of SOD2 and CSF3 modulating kinases
## 4.1 Genes: SOD2, CSF3, CTSS, NAMPT, TNIP1, CXCL2
```{r}
sod2_csf3_modulating_kinases <- genotype_by_Tcell_dose_diff_test_results %>% 
  filter(gene_short_name %in% c("SOD2","CSF3",'CTSS','NAMPT','TNIP1','CXCL2'), term != "(Intercept)", q_value < 0.05 & abs(normalized_effect) > 1) %>% 
  arrange(desc(normalized_effect)) %>% 
  group_by(gene_short_name) %>% 
  dplyr::slice(1:10) %>% 
  dplyr::select(gene_short_name,normalized_effect, term)

sod2_csf3_modulating_kinases$gene_id <- sapply(sod2_csf3_modulating_kinases$term, function(x){
  stringr::str_split(x, pattern = "_id")[[1]][2]
})

sod2_csf3_modulating_kinases <- sod2_csf3_modulating_kinases %>%
	dplyr::distinct(gene_short_name, gene_id, .keep_all = TRUE)
```

```{r}
sod2_csf3_modulating_kinases_beta_heatmap <- sod2_csf3_modulating_kinases %>%
  dplyr::select(gene_short_name, normalized_effect,gene_id) %>%
  tidyr::spread(key = gene_id, value = normalized_effect) %>%
  as.data.frame()

row.names(sod2_csf3_modulating_kinases_beta_heatmap) <- sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name
sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name <- NULL

sod2_csf3_modulating_kinases_beta_heatmap[is.na(sod2_csf3_modulating_kinases_beta_heatmap)] <- 0

hmcols <- colorRampPalette(colors = c("blue", "white","red"))(35)
paletteLength <- 35

myBreaks <- c(seq(min(sod2_csf3_modulating_kinases_beta_heatmap), 0, length.out=ceiling(paletteLength/2)+1), seq(max(sod2_csf3_modulating_kinases_beta_heatmap)/floor(paletteLength/2), max(sod2_csf3_modulating_kinases_beta_heatmap), length.out=floor(paletteLength/2)))

pheatmap::pheatmap(sod2_csf3_modulating_kinases_beta_heatmap, 
                   clustering_method = "ward.D2", 
                   cluster_rows = F, 
                   color = hmcols,
                   breaks = myBreaks,
                   file = "CRISPRI_Figures/Heatmaps/SOD2_CSF3_modulators.png",
                   fontsize = 9,
                   treeheight_col = 10,
                   height = 3, 
                   width = 6)
```

## 4.2 Genes: top T cell exposure genes
```{r}
sod2_csf3_modulating_kinases <- genotype_by_Tcell_dose_diff_test_results %>% 
  filter(gene_short_name %in% top_Tcell_dose_gene$gene_short_name, term != "(Intercept)", q_value < 0.05 & abs(normalized_effect) > 1) %>% 
  arrange(desc(normalized_effect)) %>% 
  group_by(gene_short_name) %>% 
  dplyr::slice(1:10) %>% 
  dplyr::select(gene_short_name, normalized_effect, term)

sod2_csf3_modulating_kinases$gene_id <- sapply(sod2_csf3_modulating_kinases$term, function(x){
  stringr::str_split(x, pattern = "_id")[[1]][2]
})

sod2_csf3_modulating_kinases <- sod2_csf3_modulating_kinases %>%
  dplyr::distinct(gene_short_name, gene_id, .keep_all = TRUE)
```

```{r}
sod2_csf3_modulating_kinases_beta_heatmap <- sod2_csf3_modulating_kinases %>%
  dplyr::select(gene_short_name, normalized_effect,gene_id) %>%
  tidyr::spread(key = gene_id, value = normalized_effect) %>%
  as.data.frame()

row.names(sod2_csf3_modulating_kinases_beta_heatmap) <- sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name
sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name <- NULL

sod2_csf3_modulating_kinases_beta_heatmap[is.na(sod2_csf3_modulating_kinases_beta_heatmap)] <- 0

hmcols <- colorRampPalette(colors = c("blue", "white","red"))(35)
paletteLength <- 35

myBreaks <- c(seq(min(sod2_csf3_modulating_kinases_beta_heatmap), 0, length.out=ceiling(paletteLength/2)+1), seq(max(sod2_csf3_modulating_kinases_beta_heatmap)/floor(paletteLength/2), max(sod2_csf3_modulating_kinases_beta_heatmap), length.out=floor(paletteLength/2)))

pheatmap::pheatmap(sod2_csf3_modulating_kinases_beta_heatmap, 
                   clustering_method = "ward.D2", 
                   cluster_rows = F, 
                   color = hmcols,
                   breaks = myBreaks,
                   file = "CRISPRI_Figures/Heatmaps/SOD2_CSF3_modulators_top_Tcell_dose.png",
                   fontsize = 9,
                   treeheight_col = 10,
                   height = 3, 
                   width = 8)
```

## 4.3 Genes: Pro-inflammatory cytokines
```{r}
Pro_inf <- c("IL1A", "IL1B", "IL17A", "IL17F", "IL18", "TNF", "CSF2", "IFNG", "CCL2", "CCL3", "CCL4", "CCL5", "CCL11", "CCL13", "CXCL10", "CXCL8", "IL12A", "IL12B", "IL4", "IL10", "IL13", "IL1RN", "TGFB1", "TGFB2", "TGFB3")
```

```{r}
sod2_csf3_modulating_kinases <- genotype_by_Tcell_dose_diff_test_results %>%
  filter(gene_short_name %in% Pro_inf, term != "(Intercept)", q_value < 0.05 & abs(normalized_effect) > 1) %>% 
  arrange(desc(normalized_effect)) %>%
  group_by(gene_short_name) %>%
  dplyr::slice(1:10) %>%
  dplyr::select(gene_short_name, normalized_effect, term)

sod2_csf3_modulating_kinases$gene_id <- sapply(sod2_csf3_modulating_kinases$term, function(x){
  stringr::str_split(x, pattern = "_id")[[1]][2]
})

sod2_csf3_modulating_kinases <- sod2_csf3_modulating_kinases %>%
  dplyr::distinct(gene_short_name, gene_id, .keep_all = TRUE)
```

```{r}
sod2_csf3_modulating_kinases_beta_heatmap <- sod2_csf3_modulating_kinases %>%
  dplyr::select(gene_short_name, normalized_effect, gene_id) %>%
  tidyr::spread(key = gene_id, value = normalized_effect) %>%
  as.data.frame()

row.names(sod2_csf3_modulating_kinases_beta_heatmap) <- sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name
sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name <- NULL

sod2_csf3_modulating_kinases_beta_heatmap[is.na(sod2_csf3_modulating_kinases_beta_heatmap)] <- 0

hmcols <- colorRampPalette(colors = c("white","red"))(18)
paletteLength <- 18
myBreaks <- c(seq(0, max(sod2_csf3_modulating_kinases_beta_heatmap), length.out=paletteLength))

pheatmap::pheatmap(sod2_csf3_modulating_kinases_beta_heatmap,
                   clustering_method = "ward.D2",
                   cluster_rows = F,
                   color = hmcols,
                   breaks = myBreaks,
                   file = "CRISPRI_Figures/Heatmaps/Pro_inf_modulators.png",
                   fontsize = 9,
                   treeheight_col = 10,
                   height = 3,
                   width = 6)
```

## 4.4 Genes: top T cell exposure genes + CD274, IDO1
```{r}
sod2_csf3_modulating_kinases <- genotype_by_Tcell_dose_diff_test_results %>%
  filter(gene_short_name %in% c(top_Tcell_dose_gene$gene_short_name, 'CD274','IDO1'), term != "(Intercept)", q_value < 0.05 & abs(normalized_effect) > 1) %>% 
  arrange(desc(normalized_effect)) %>% 
  group_by(gene_short_name) %>% 
  dplyr::slice(1:10) %>% 
  dplyr::select(gene_short_name,normalized_effect, term)

sod2_csf3_modulating_kinases$gene_id <- sapply(sod2_csf3_modulating_kinases$term, function(x){
  stringr::str_split(x, pattern = "_id")[[1]][2]
})

sod2_csf3_modulating_kinases <- sod2_csf3_modulating_kinases %>%
  dplyr::distinct(gene_short_name, gene_id, .keep_all = TRUE)
```

```{r}
sod2_csf3_modulating_kinases_beta_heatmap <- sod2_csf3_modulating_kinases %>%
  dplyr::select(gene_short_name, normalized_effect, gene_id) %>%
  tidyr::spread(key = gene_id, value = normalized_effect) %>%
  as.data.frame()

row.names(sod2_csf3_modulating_kinases_beta_heatmap) <- sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name
sod2_csf3_modulating_kinases_beta_heatmap$gene_short_name <- NULL

sod2_csf3_modulating_kinases_beta_heatmap[is.na(sod2_csf3_modulating_kinases_beta_heatmap)] <- 0

hmcols <- colorRampPalette(colors = c("blue", "white","red"))(35)
paletteLength <- 35
myBreaks <- c(seq(min(sod2_csf3_modulating_kinases_beta_heatmap), 0, length.out=ceiling(paletteLength/2)+1), seq(max(sod2_csf3_modulating_kinases_beta_heatmap)/floor(paletteLength/2), max(sod2_csf3_modulating_kinases_beta_heatmap), length.out=floor(paletteLength/2)))

pheatmap::pheatmap(sod2_csf3_modulating_kinases_beta_heatmap,
                   clustering_method = "ward.D2",
                   cluster_rows = F,
                   color = hmcols,
                   breaks = myBreaks,
                   file = "CRISPRI_Figures/Heatmaps/SOD2_CSF3_modulators_top_Tcell_dose_CD274_IDO1.png",
                   fontsize = 9,
                   treeheight_col = 10,
                   height = 3,
                   width = 8)
```

# 5.Kinases that modulate 
## 5.1 Select kinased based on dose analysis
```{r}
intersection_counts <- numeric(length(kinases))
names(intersection_counts) <- kinases  # Name the vector with kinase names
```

```{r}
for (target in kinases) {
  subset_results <- Tcell_genotype_diff_results[
    Tcell_genotype_diff_results$gene_id == target &
    Tcell_genotype_diff_results$term == "dose" &
    Tcell_genotype_diff_results$q_value < 0.05]
  intersection_counts[target] <- length(intersect(Tcell_dose_diff_genes, subset_results$id))
}

result_df <- data.frame(
  Kinase = names(intersection_counts),
  IntersectionCount = intersection_counts
)
sorted_result <- result_df[order(-result_df$IntersectionCount), ]
```

```{r}
print("Sorted kinases based on dose analysis intersection")
print(head(sorted_result, 100))
```

## 5.2 Select kinased based on if T cell dose response genes are being perturbed in untreated
```{r}
intersection_counts <- numeric(length(kinases))
names(intersection_counts) <- kinases
```

```{r}
for (target in kinases) {
  subset_results <- untreated_genotype_diff_results[
    untreated_genotype_diff_results$gene_id == target &
    untreated_genotype_diff_results$term != "(Intercept)" &
    untreated_genotype_diff_results$q_value < 0.01]
  
  intersection_counts[target] <- length(intersect(Tcell_dose_diff_genes, subset_results$id))
}

result_df <- data.frame(
  Kinase = names(intersection_counts),
  IntersectionCount = intersection_counts
)
sorted_result <- result_df[order(-result_df$IntersectionCount), ]
```

```{r}
print("Sorted kinases based on untreated perturbation of T-cell dose genes")
print(head(sorted_result, 100))
```

```{r}
write.csv(sorted_result, "deg_ntc_kinases.csv", row.names = FALSE)
write.csv(Tcell_dose_diff_test, "Tcell_dose_diff_test.csv", row.names = FALSE)
```
